---
title: "RNAseq vs qPCR comparison"
author: "Sam Haynes"
date: "20/08/2021"
output: pdf_document
---

Now we have QuantSeq and 5PSeq datasets we can compare the abundance of transcripts in these datasets to that determined by qPCR.

Constructs in QuantSeq (Standard RNA-Seq but PolyA Anchored)

Sample Name | Construct Name
------------|---------------
A1 B1       |pRPS3-mCherry-tRPS3_WT
A2 B2 E1 E2 |pRPS3-mCherry-tRPS3_mod0
A3 B3       |pRPS3-mCherry-tRPS3_modB
A4 B4       |pRPS3-mCherry-tRPS3_modE
A5 B5       |pTSA1-mCherry-tTSA1_WT
A6 B6       |pTSA1-mCherry-tTSA1_mod0
A7 B7       |pTSA1-mCherry-tTSA1_modB
A8 B8       |pTSA1-mCherry-tTSA1_modE
A9 B9       |POT1-ccdB
A10 B10     |pRPS3-mCherry-tRPS3_modA
A11 B11     |pTSA1-mCherry-tTSA1_modA
A12 B12     |pPGK1-mCherry-tRPS3_WT
A13 B13     |pPGK1-mCherry-tRPS3_mod0
A14 B14     |pPGK1-mCherry-tRPS3_modA
A15 B15     |pPGK1-mCherry-tRPS3_modB
A16 B16     |pPGK1-mCherry-tRPS3_modE

Constructs in 5PSeq (RNA-Seq of transcripts with 5'Cap removed)

Sample Name | Construct Name
------------|---------------
1dt 2dt     |POT1-ccdB
3dt 4dt     |pRPS3-mCherry-tRPS3_WT
5dt 6dt     |pRPS3-mCherry-tRPS3_mod0
7dt 8dt     |pRPS3-mCherry-tRPS3_modA
9dt 10dt    |pRPS3-mCherry-tRPS3_modB
11dt 12dt   |pRPS3-mCherry-tRPS3_modC
13dt 14dt   |pRPS3-mCherry-tRPS3_modD
15dt 16dt   |pRPS3-mCherry-tRPS3_modE

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(DESeq2)
source("/home/dimmest-p/Documents/PhD/3UTR_motifs/motifPrediction/chimera_project_manuscript/raw_data_analysis/code/shared_figure_formatting.R")
```

```{r functions}

read_in_RNAseq <- function(filepath){
  # read in only name and counts
  RNAseq_tibble <- read_tsv(filepath, comment = "#") %>%
    select(1,7)
  
  colnames(RNAseq_tibble) <- c("Geneid", "count")
  
  # replace specific construct Geneid with generic Geneid to allow comparison
  
  construct_URA3_counts <- RNAseq_tibble %>% 
    filter(str_detect(Geneid, "URA3"))
  
  new_name_construct_URA3_counts <- tibble(Geneid = "plasmid_URA3_construct", "count" = construct_URA3_counts %>% pull(2))
  
  construct_RNAseq_tibble <- RNAseq_tibble  %>%
    rows_delete(construct_URA3_counts) %>%
    rows_insert(new_name_construct_URA3_counts)
  
  if(sum(str_detect(RNAseq_tibble$Geneid, "mCh")) > 0){
    construct_counts <- RNAseq_tibble %>% 
    filter(str_detect(Geneid, "mCh"), !str_detect(Geneid, "URA3"))
    
    new_name_construct_counts <- tibble(Geneid = "plasmid_construct", 
                                        "count" = construct_counts %>% pull(2))
    construct_RNAseq_tibble <- construct_RNAseq_tibble %>%
    rows_delete(construct_counts) %>%
    rows_insert(new_name_construct_counts)
  }
  
  # find sample name and add as colname
  colnames(construct_RNAseq_tibble) <- c("Geneid", str_extract(filepath, "(?<=/)[ABEdt0-9]{2,4}(?=/)"))
  
  construct_RNAseq_tibble
}

sample_to_construct_dictionary <- function(filename){
  read_csv(here(filename)) %>%
    mutate(promoter = str_extract(construct_name, "^p[A-Z0-3]+")) %>%
    mutate(terminator = str_extract(construct_name, "t[A-Z0-3]+(?=_)")) %>%
    mutate(mod_label = str_extract(construct_name, "(?<=_)[a-zA-Z0-3]+(?=_)")) %>%
    select(-construct_name) 
}

```

```{r read-RNAseq-counts}
all_QuantSeq_count_files <- list.files(path = here("data/output/counts/QuantSeq"), pattern = ".txt$", recursive = TRUE, full.names = TRUE)

all_5PSeq_count_files <- list.files(path = here("data/output/counts/5PSeq"), pattern = ".txt$", recursive = TRUE, full.names = TRUE)

full_quantseq_count <- all_QuantSeq_count_files %>%
  lapply(read_in_RNAseq) %>%
  purrr::reduce(full_join, by = "Geneid")

full_5PSeq_count <- all_5PSeq_count_files %>%
  lapply(read_in_RNAseq) %>%
  purrr::reduce(full_join, by = "Geneid")

full_quantseq_count_matrix <- full_quantseq_count %>% 
  select(-Geneid) %>%
  as.matrix()

# replace NAs with 0 
full_quantseq_count_matrix[is.na(full_quantseq_count_matrix)] <- 0

rownames(full_quantseq_count_matrix) <- full_quantseq_count %>% 
  pull(Geneid)

```

```{r sample-name-to-contruct}
sample_to_construct_dictionary_QuantSeq <- sample_to_construct_dictionary("./data/input/QuantSeq_sample_to_construct_table.csv")

sample_to_construct_dictionary_5PSeq <- sample_to_construct_dictionary("./data/input/5PSeq_sample_to_construct_table.csv")

#  swap POT1 empty vector values from NA to "POT"
sample_to_construct_dictionary_QuantSeq <- sample_to_construct_dictionary_QuantSeq %>%
    rows_update(tibble(sample_name = c("A9", "B9"), promoter = "POT1", terminator = "POT1", mod_label = "POT1")) %>%
  mutate(mod_label = factor(mod_label,
                       levels = c(construct_to_label_dictionary_TSA1_RPS3$construct, "POT1"),
                       labels = c(construct_to_label_dictionary_TSA1_RPS3$label, "POT1")))

sample_to_construct_dictionary_5PSeq <- sample_to_construct_dictionary_5PSeq %>%
    rows_update(tibble(sample_name = c("1dt", "2dt"), promoter = "POT1", terminator = "POT1", mod_label = "POT1")) %>%
  mutate(mod_label = factor(mod_label,
                       levels = c(construct_to_label_dictionary_TSA1_RPS3$construct, "POT1"),
                       labels = c(construct_to_label_dictionary_TSA1_RPS3$label, "POT1")))
```

```{r normalise-to-URA3}
URA3_norm_counts_QuantSeq <- full_quantseq_count %>%
  pivot_longer(-Geneid, names_to = "sample_name", values_to = "counts") %>%
  inner_join(sample_to_construct_dictionary_QuantSeq) %>%
  filter(str_detect(Geneid,"plasmid"), terminator != "POT1") %>%
  pivot_wider(names_from = "Geneid", values_from = "counts") %>%
  mutate(ratio = plasmid_construct / plasmid_URA3_construct)

URA3_norm_counts_5PSeq <- full_5PSeq_count %>%
  pivot_longer(-Geneid, names_to = "sample_name", values_to = "counts") %>%
  inner_join(sample_to_construct_dictionary_5PSeq) %>%
  filter(str_detect(Geneid,"plasmid"), terminator != "POT1") %>%
  pivot_wider(names_from = "Geneid", values_from = "counts") %>%
  mutate(ratio = plasmid_construct / plasmid_URA3_construct)

ggplot(URA3_norm_counts_QuantSeq) +
  geom_point(aes(y=ratio,x=mod_label)) +
  facet_wrap(~terminator+promoter)

ggplot(URA3_norm_counts_5PSeq) +
  geom_point(aes(y=ratio,x=mod_label)) +
  facet_wrap(~terminator+promoter)
```

```{r log2-fold-change}
promoter_mean_mod_NNN_QuantSeq <- URA3_norm_counts_QuantSeq %>% 
  filter(mod_label == "mod_NNN") %>%
  group_by(promoter) %>%
  summarise(mean_mod_NNN = mean(ratio))

construct_log2_fold_change_QuantSeq <- URA3_norm_counts_QuantSeq %>%
  inner_join(promoter_mean_mod_NNN_QuantSeq) %>%
  transmute(promoter,terminator,mod_label,rel_abund_delta_deltacq = ratio / mean_mod_NNN)

construct_plot_QuantSeq <- ggplot(construct_log2_fold_change_QuantSeq) +
  RNA_relative_abundance_figure_options +
  facet_wrap(~terminator+promoter) +
  labs(x ="Fold change in RNA abundance relative to mod_NNN", y = "", title = "QuantSeq")

promoter_mean_mod_NNN_5PSeq <- URA3_norm_counts_5PSeq %>% 
  filter(mod_label == "mod_NNN") %>%
  group_by(promoter) %>%
  summarise(mean_mod_NNN = mean(ratio))

construct_log2_fold_change_5PSeq <- URA3_norm_counts_5PSeq %>%
  inner_join(promoter_mean_mod_NNN_5PSeq) %>%
  transmute(promoter,terminator,mod_label,rel_abund_delta_deltacq = ratio / mean_mod_NNN)

construct_plot_5PSeq <- ggplot(construct_log2_fold_change_5PSeq) +
  RNA_relative_abundance_figure_options +
  facet_wrap(~terminator+promoter) +
  labs(x ="Fold change in RNA abundance\n relative to mod_NNN", y = "", title = "5PSeq")

construct_plot <- cowplot::plot_grid(construct_plot_QuantSeq,construct_plot_5PSeq, rel_widths = c(2,1))

ggsave(here("RNASeq_construct_plot.RData"), construct_plot, height = 4, width= 4)
```

```{r import_qPCR_results}
TSA1_deltadeltacq <- read_csv("/home/dimmest-p/Documents/PhD/3UTR_motifs/motifPrediction/chimera_project_manuscript/raw_data_analysis/data/norm_qpcr/motif_context_dependence/pTSA1_pPGK1_pSRO9_tTSA1_deltadeltacq_platesnorm_summarise.csv") %>%
  mutate(mod_label = factor(construct,
                       levels = construct_to_label_dictionary_TSA1_RPS3$construct,
                       labels = construct_to_label_dictionary_TSA1_RPS3$label),
         promoter = factor(promoter, levels = c("pTSA1","pPGK1","pSRO9")))

RPS3_deltadeltacq <- read_csv("/home/dimmest-p/Documents/PhD/3UTR_motifs/motifPrediction/chimera_project_manuscript/raw_data_analysis/data/norm_qpcr/motif_context_dependence/pRPS3_pPGK1_pSRO9_tRPS3_deltadeltacq_platesnorm_summarise.csv") %>%
  mutate(mod_label = factor(construct,
                       levels = construct_to_label_dictionary_TSA1_RPS3$construct,
                       labels = construct_to_label_dictionary_TSA1_RPS3$label),
         promoter = factor(promoter, levels = c("pRPS3","pPGK1","pSRO9")))

TSA1_deltadeltacq_plot <- ggplot(data = TSA1_deltadeltacq)+
  RNA_relative_abundance_figure_options +
  scale_colour_manual(values=RPS3_TSA1_colour_scheme) +
  theme(axis.text.y=element_text(colour=RPS3_TSA1_colour_scheme, face = "bold")) +
  labs(x ="Fold change in RNA abundance relative to tTSA1 mod_NNN", y = "") +
  facet_wrap(~promoter,ncol = 3) 

RPS3_deltadeltacq_plot <- ggplot(data = RPS3_deltadeltacq) +
  RNA_relative_abundance_figure_options +
  scale_colour_manual(values=RPS3_TSA1_colour_scheme) +
  theme(axis.text.y=element_text(colour=RPS3_TSA1_colour_scheme, face = "bold")) +
  labs(x ="Fold change in RNA abundance relative to tRPS3 mod_NNN", y = "") +
  facet_wrap(~promoter,ncol = 3) 

TSA1_deltadeltacq_plot

RPS3_deltadeltacq_plot

```

```{r compare_qPCR_and_quantseq}
mean_qpcr_vs_QuantSeq_dataset <- construct_log2_fold_change_QuantSeq %>%
              mutate(source = "RNA-Seq", type = "QuantSeq") %>%
  bind_rows(bind_rows(TSA1_deltadeltacq %>% filter(promoter == "pTSA1"),
                       RPS3_deltadeltacq%>% filter(promoter %in% c("pRPS3",
                                                                   "pPGK1"))) %>% 
              select(promoter, terminator, mod_label, rel_abund_delta_deltacq) %>%
              mutate(source = "qPCR", type = "QuantSeq") %>%
              filter(mod_label %in% c("mod_NNN","mod_NAA","WT", 
                                      "mod_HTH","mod_NTN"))) %>%
  group_by(promoter, source, mod_label, terminator, type) %>%
  summarise(mean_rel_abund = mean(rel_abund_delta_deltacq)) %>%
  ungroup() %>%
  pivot_wider(names_from = "source", values_from = "mean_rel_abund")

mean_qpcr_vs_5PSeq_dataset <- construct_log2_fold_change_5PSeq %>%
  filter(is.finite(rel_abund_delta_deltacq)) %>%
  mutate(source = "RNA-Seq", type = "5PSeq") %>%
  bind_rows(RPS3_deltadeltacq%>% filter(promoter == "pRPS3") %>% 
              select(promoter, terminator, mod_label, rel_abund_delta_deltacq) %>%
              mutate(source = "qPCR", type = "5PSeq")) %>%
  group_by(promoter, source, mod_label, terminator, type) %>%
  summarise(mean_rel_abund = mean(rel_abund_delta_deltacq)) %>%
  ungroup() %>%
  pivot_wider(names_from = "source", values_from = "mean_rel_abund")
   
qPCR_vs_RNASeq_plot <- ggplot(mean_qpcr_vs_QuantSeq_dataset %>% bind_rows(mean_qpcr_vs_5PSeq_dataset)) +
  geom_point(aes(x = `RNA-Seq`, y = qPCR, colour = promoter, shape = terminator)) +
  geom_abline(slope = 1, intercept = 0) +
  facet_wrap(~type)

save(qPCR_vs_RNASeq_plot, file = here("qPCR_vs_RNASeq_plot.Rdata"))
```

# Could normalising to URA3 cause strange pPGK1 behaviours? Check norm to DESeq2.

```{r create-DESeq2-input}
# tibble holding what condition each column in the count matrix correlates to
coldata <- tibble(sample_name = colnames(full_quantseq_count_matrix)) %>%
  inner_join(sample_to_construct_dictionary) %>% 
  mutate(col = 1:34) %>%
  filter(terminator == "tRPS3") %>%
  mutate(promoter = as.factor(promoter), 
         terminator = as.factor(terminator), 
         mod_label = as.factor(mod_label)) %>%
  column_to_rownames("sample_name")

deseq_count_set <-  DESeqDataSetFromMatrix(countData = full_quantseq_count_matrix[,coldata %>% pull(col)],
                                           colData = coldata,
                                           design = ~ promoter * mod_label)
```

```{r run-DESeq}
dds <- DESeq(deseq_count_set)
```

```{r view_results}
res <- results(dds, contrast = c("promoter:mod_label", "pRPS3:modA", "pPGK1:modA"))
res[rownames(res) == "plasmid_construct",]
rownames(res[res[!is.na(res$padj),]$padj < 0.05,])
plotMA(res, ylim=c(-2,2))
```

```{r pPGK1-behaviour}

deseq_scale_factor <- tibble(sample_name = colnames(full_quantseq_count_matrix[,coldata %>% pull(col)]), 
                    DESeq_norm = estimateSizeFactorsForMatrix(full_quantseq_count_matrix[, coldata %>% pull(col)]))

URA3_counts_vs_DESeq_norm <- URA3_norm_counts %>% 
  select(sample_name, plasmid_URA3_construct) %>% 
  inner_join(deseq_scale_factor) %>% 
  filter(sample_name %in% c("A12","B12","A13","B13","A14",
                            "B14","A15","B15","A16","B16"))

deseq_norm_counts <- full_quantseq_count %>%
  pivot_longer(-Geneid, names_to = "sample_name", values_to = "counts") %>%
  inner_join(sample_to_construct_dictionary) %>%
  inner_join(deseq_scale_factor) %>%
  filter(str_detect(Geneid,"plasmid_construct"), terminator != "POT1") %>%
  mutate(norm_counts = counts / DESeq_norm)

ggplot(deseq_norm_counts) +
  geom_point(aes(y=norm_counts,x=mod_label)) +
  facet_wrap(~terminator+promoter)
```